Syntax-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-05-14 paper/paper.tex
various additions

Nevertheless, lexical exactness is, in this context, just as important as grammaticality. 
An important part of the design of the Controlled Natural Language (CNL) the grammar in such a system describes becomes, then, the creation of a translation lexicon. 
In many cases, this is done for the most part manually, resulting in a time consuming task requiring significant linguistic knowledge. 
When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}. 
None of them is, however, suitable for the common case in which only a limited number of example data is available.

@@

This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{ws-2019-universal} and Grammatical Framework (GF) \cite{ranta-etal-2020-abstract}.

While both formalisms, the former a dependency and the latter a constituency grammar\todo{too quick, say more about both}, independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing. 
As a consequence, several robust dependency parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.

UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL). \todo{say what the RGL is} 
This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.

Concretely, the system we propose requires the following elements, whose reciprocal relations are shown in Figure \ref{overview}: \smallskip

@@

% TODO: replace CE with CA and omit CP entirely?
\begin{figure}[h]
  \centering
  \includegraphics[width=0.45\textwidth]{figures/overview.png}
  \caption[System overview]{System overview. Parts in grey are currently at the prototype stage and will not be further discussed in this paper.} \label{elems}
  \label{overview}
\end{figure}

@@

As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match any of the alignment criteria. 
If this is the case, they are added to a collection of alignments, which are also represented as a pair of UD trees associated with some metadata. \todo{explain what metadata?}
Such collection is what the function will return after aligning all the dependency subtrees. % TODO: explain or omit metadata!

Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection. This is important... \todo{explain why}% TODO:

The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.  % TODO: explain & motivate sorting -> maybe say more about UD before?

@@

\subsection{Generating grammar rules}
The CE module described so far outputs pairs of UD trees. 
While converting them to GF Abstract Syntax Trees (ASTs) is \texttt{gf-ud}'s core functionality, to build a compilable GF translation lexicon it is necessary to turn such trees into grammar rules.

@@

\section{Evaluation} \label{evaluation}
% TODO: move around things to have something here (or outline structure of the section)
\subsection{Data}
Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out using manually annotated data. 
To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of 1000 handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.}\todo{not even, I fear. Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2018 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{zeman-etal-2018-conll}.
PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
Using this data limits the amount of alignment errors that are due annotation issues to a minimum, even though a small number of inconsistencies and imprecisions is present even in this corpus.

When it comes to testing the program on raw text, we use two bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish). 
For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{thesis}.
When using raw text, our parser of choice is UDPipe \cite{straka-etal-2016-udpipe}. 
In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus.

@@

\subsection{Evaluating CE} %TODO: better title?
While in this work we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which our system can be put to use. 
For instance, it is easy to imagine using it to improve the UI of online translation memories or to facilitate reading parallel texts in the context of language learning.
For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the subsequent stages of our proposed lexicon generation pipeline.

@@

Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files excepts with regards to tokenization. 
On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison. 
In both cases, \texttt{fast\_align} is run with the recommended parameters.

@@

The results of this comparison are summarized in Table \ref{raw_fast}\todo{position properly; run experiments (this is new) and add actual results}.

@@

\subsection{MT experiments}
The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 
Because we are focussing on using CA in the context of domain-specific MT, in this case we do not make use of the PUD treebanks, where sentences come from a variety of different sources, but rather of the two course plans corpora. 
For this evaluation, we do not construct a grammar specific to the course plans domain: it is sufficient to extend the extraction grammar itself with preexisting syntax rules defined in the RGL, so to allow for more variation.

>>>>

Nevertheless, lexical exactness is, in this context, just as important as grammaticality. 
An important part of the design of the Controlled Natural Language (CNL) the grammar in such a system describes becomes, then, the creation of a translation lexicon. 
In many cases, this is done for the most part manually, resulting in a time consuming task requiring significant linguistic knowledge. 
When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}. 
None of them is, however, suitable for the common case in which only a limited amount of example data is available.

@@

This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-etal-2020-abstract} and Universal Dependencies (UD) \cite{ws-2019-universal}.

GF is a constituency grammar formalisms and programming language designed with the parallel between programming and natural languages in mind: grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages. 
In the case of translation, strings in the source language are \textit{parsed} to Abstract Syntax Trees (ASTs) which are then \textit{linearized} to target language strings.

UD, on the other hand, is a dependency grammar formalism meant for cross-linguistically consistent grammatical annotation.
As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each \textit{head} is simply put in relation with the ones that depend on it, called its \textit{dependents}, via directed labelled links, specifying the syntactic relation between them, and the root element of a sentence is attached to a fake node with a link labelled \texttt{root}.
Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word of a sentence, but also on its POS tag, lemma and other morphological features.

While both formalisms independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing. 
As a consequence, several robust dependency parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.

UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL), which covers the morphology and basic syntax of over 30 languages.
This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.

Concretely, the system we propose consists of the following elements: \smallskip

@@

@@

As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match any of the alignment criteria. 
If this is the case, they are added to a collection of alignments, which are also represented as a pair of UD trees associated with some metadata, such as the id of the sentence they were extracted from.
Such collection is what the function will return after aligning all the dependency subtrees.

Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection. This is important not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.

The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.
Subtrees are sorted based on their dependency label so to give higher priority to pairs with the same root label.

@@

\subsection{Generating grammar rules}
The CE module described so far outputs pairs of UD trees. 
While converting them to GF ASTs is \texttt{gf-ud}'s core functionality, to build a compilable GF translation lexicon it is necessary to turn such trees into grammar rules.

@@

\section{Evaluation} \label{evaluation}
In this section, we evaluate the system proposed in the above. 
We first introduce the data used for such evaluation. 
After that, we describe our experiments, aimed at putting both the CE module \textit{per se} and the entire system from parsing to lexicon generation to the test, and present our results.
% TODO: move around things to have something here (or outline structure of the section)
\subsection{Data} \label{data}
Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out using manually annotated data. 
To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of 1000 handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.}\todo{not even, I fear. Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2018 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{zeman-etal-2018-conll}.
PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
Using this data limits the amount of alignment errors that are due annotation issues to a minimum, even though a small number of inconsistencies and imprecisions is present even in this corpus.

When it comes to testing the program on raw text, we use two small ($< 1000$ sentences) bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science (DMI) of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish). 
For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{thesis}.
When using raw text, our parser of choice is UDPipe \cite{straka-etal-2016-udpipe}. 
In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus.

@@

\subsection{Evaluating CE} \label{ceval}
While in this work we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which our system can be put to use. 
For instance, it is easy to imagine using it to improve the UI of online translation memories or to facilitate reading parallel texts in the context of language learning.
For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the subsequent stages of our proposed lexicon generation pipeline.

@@

Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files excepts with regards to tokenization. 
On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.

In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured to only extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.

@@

The results of this comparison are summarized and compared with those obtained on the sample of PUD treebanks in Table \ref{raw_fast}.

What is immediately evident, but not unexpected, is a decrease in precision. 
The percentage of correct alignments, however, stays significantly higher than that obtained with \texttt{fast\_align} in the previous experiment, even when the model was trained on the full PUD corpus: even though percentages seem similar for English-Swedish, it must be taken into account that the CSE corpus is roughly half the size of the PUD corpus.

Results are less encouraging in terms of recall: the number of alignments extracted from the course plans corpora is similar to that obtained from the PUD treebank sample, despite the difference in size. 
This is explained in part by the presence, in the course plans corpora, of a large amount of very short sentences, and in part by the fact that parse errors introduced by UDPipe make it impossible to align many subsentences without a significant loss in terms of precision.

@@

\subsection{MT experiments} \label{mtexp}
The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 
Because we are focussing on using CA in the context of domain-specific MT, in this case we do not make use of the PUD treebanks, where sentences come from a variety of different sources, but rather of the two course plans corpora. 
For this evaluation, we do not construct a grammar specific to the course plans domain: it is sufficient to extend the extraction grammar itself with preexisting syntax rules defined in the RGL, so to allow for more variation.
